---
title: "Navigating the ATC Hierarchy"
author: "Lucas VHH TRAN"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    includes:
      in_header: header.html
vignette: >
  %\VignetteIndexEntry{Navigating the ATC Hierarchy}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width  = 8,
  fig.height = 5,
  message    = FALSE,
  warning    = FALSE
)
```

<!-- ── Branded banner ───────────────────────────────────────────── -->
<header style="
  background: linear-gradient(135deg, #05161a 0%, #0B525B 55%, #168aad 100%);
  color: #fff;
  padding: 1.5em 2.2em 1.3em;
  margin: 0 -2em 2.2em;
  border-radius: 0 0 14px 14px;
  display: flex;
  align-items: center;
  gap: 1.1em;
  box-shadow: 0 4px 18px rgba(5,22,26,.35);
">
  <div style="font-size: 2.6em; line-height:1; filter: drop-shadow(0 2px 6px rgba(0,0,0,.4));">🧬</div>
  <div>
    <div style="font-size:1.45em; font-weight:800; letter-spacing:.04em; line-height:1;">
      atcddd
    </div>
    <div style="font-size:.8em; opacity:.80; margin-top:.18em; letter-spacing:.01em;">
      WHO ATC/DDD Drug Classification Toolkit &nbsp;·&nbsp; v0.2.0
    </div>
  </div>
  <nav style="margin-left:auto; font-size:.78em; opacity:.78; text-align:right; line-height:1.7;">
    <a href="https://www.whocc.no/atc_ddd_index/" target="_blank"
       style="color:#99e8c0; text-decoration:none; border-bottom:1px solid transparent;
              transition:border-color .18s;"
       onmouseover="this.style.borderBottomColor='#99e8c0'"
       onmouseout="this.style.borderBottomColor='transparent'">WHO ATC/DDD Index</a><br>
    Pharmacoepidemiology · Drug Utilisation Research
  </nav>
</header>

```{r include-lightbox, echo=FALSE, results='asis'}
if (file.exists('lightbox.html')) cat(readLines('lightbox.html'), sep = '\n')
```

# Introduction

The ATC hierarchy is a five-level tree that classifies every drug from broad
anatomical group down to individual chemical substance. Navigating this tree
is essential for pharmacoepidemiology — you need to answer questions like:

- *"Which drugs belong to the beta-blocker class?"*
- *"What are the children of the statin subclass?"*
- *"How many Level 5 substances are in the nervous system group?"*

This vignette shows you how to traverse, filter, and visualise the hierarchy
using `atcddd`.

```{r load_package}
library(atcddd)
library(dplyr)
library(ggplot2)
library(tidyr)
```

---

# Loading the Offline Database

We start with the bundled WHO snapshot so everything works offline:

```{r load_data}
codes_path <- system.file("extdata", "WHO_ATC_codes_2026-07-14.csv",
                          package = "atcddd")
ddd_path   <- system.file("extdata", "WHO_ATC_DDD_2026-07-14.csv",
                          package = "atcddd")

atc_codes <- readr::read_csv(codes_path, show_col_types = FALSE)
atc_ddd   <- readr::read_csv(ddd_path, show_col_types = FALSE)

# Add level and parent columns using exported functions
atc_codes <- atc_codes %>%
  mutate(
    level       = atc_level(atc_code),
    parent_code = atc_parent(atc_code)
  )

cat(sprintf("Loaded %d codes across 5 levels", nrow(atc_codes)))
```

---

# Understanding ATC Levels

The five levels form a strict tree:

```
Level 1:  C                   (1 char)      Cardiovascular system
Level 2:  C10                 (3 chars)     Lipid modifying agents
Level 3:  C10A                (4 chars)     Lipid modifying agents, plain
Level 4:  C10AA               (5 chars)     HMG CoA reductase inhibitors
Level 5:  C10AA01             (7 chars)     simvastatin
```

## Code Distribution by Level

```{r level_distribution}
atc_codes %>%
  count(level) %>%
  mutate(
    pct   = round(100 * n / sum(n), 1),
    label = case_when(
      level == 1 ~ "Anatomical main group",
      level == 2 ~ "Therapeutic subgroup",
      level == 3 ~ "Pharmacological subgroup",
      level == 4 ~ "Chemical subgroup",
      level == 5 ~ "Chemical substance"
    )
  ) %>%
  arrange(level) %>%
  select(label, n, pct)
```

Level 5 (individual substances) dominates — each anatomical group fans out
into hundreds of specific drugs.

---

# Finding Drugs by Therapeutic Class

## All Statins (HMG CoA Reductase Inhibitors, C10AA)

The most direct way to find a drug class is by its Level 4 code:

```{r find_statins}
atc_codes %>%
  filter(grepl("^C10AA", atc_code)) %>%
  select(atc_code, atc_name, level)
```

## All Beta-Blocking Agents (C07)

```{r find_beta_blockers}
# Get all codes under C07
beta_blockers <- atc_codes %>%
  filter(grepl("^C07", atc_code))

# Show only Level 4 subgroups (chemical classes)
beta_blockers %>%
  filter(level == 4) %>%
  select(atc_code, atc_name)
```

## All Level 5 Substances Under a Parent

The new `atc_descendants()` function returns every code below a given parent:

```{r substances_under_c10aa}
# Every descendant under the statin chemical class (C10AA)
atc_descendants("C10AA", atc_codes)
```

You can also limit the depth with `max_level`:

```{r descendants_limited}
# Under cardiovascular (C) but only down to Level 3
atc_descendants("C", atc_codes, max_level = 3)
```

---

# Parent-Child Relationships

The exported functions `atc_level()` and `atc_parent()` let you reconstruct
the tree:

```{r parent_child}
# Every code and its direct parent
atc_codes %>%
  filter(level >= 2) %>%
  select(atc_code, atc_name, level, parent_code) %>%
  head(10)
```

## Find All Children of a Specific Code

```{r children_of}
# Use the exported atc_children() function
atc_children("C10AA", atc_codes)
```

## Find the Full Ancestry of a Drug

```{r ancestry}
trace_ancestry <- function(code) {
  parts <- list()
  current <- code
  while (!is.na(current) && current != "") {
    row <- atc_codes %>% filter(atc_code == current)
    if (nrow(row) == 0) break
    parts[[length(parts) + 1]] <- row
    current <- atc_parent(current)
  }
  bind_rows(rev(parts)) %>%
    select(atc_code, atc_name, level)
}

# Trace paracetamol from Level 1 down
trace_ancestry("N02BE01")
```

This ancestry trace is valuable for:
- **Study inclusion criteria** — "all analgesics except opioids"
- **Pharmacovigilance** — signal detection within a therapeutic class
- **Drug utilisation** — aggregating consumption by class

---

# Working with the Live API

For up-to-date data, use `get_atc_hierarchy()`:

```{r live_api, eval=FALSE}
# Get the complete nervous-system hierarchy (live)
ns_tree <- get_atc_hierarchy("N", max_levels = 5)

# This returns the same enriched structure
ns_tree %>%
  select(atc_code, atc_name, level, parent_code, has_children) %>%
  head(10)
```

`get_atc_hierarchy()` adds a `has_children` column — `TRUE` if the code has
any sub-classifications, which is useful for building interactive tree widgets.

---

# Comparing Therapeutic Classes

## How Many Drugs in Each Group?

```{r group_sizes}
# Count Level 5 substances per Level 1 anatomical group
atc_codes %>%
  filter(level == 5) %>%
  mutate(anatomical_group = substr(atc_code, 1, 1)) %>%
  count(anatomical_group, sort = TRUE) %>%
  rename(n_substances = n)
```

Groups J (antiinfectives), N (nervous system), and A (alimentary tract) are the
largest — reflecting decades of drug development in those therapeutic areas.

---

# Visualising the Hierarchy

## Treemap of ATC Level 1 Groups

```{r treemap, fig.width=10, fig.height=6}
atc_codes %>%
  filter(level == 5) %>%
  mutate(group = substr(atc_code, 1, 1)) %>%
  count(group, sort = TRUE) %>%
  mutate(
    group_label = case_when(
      group == "A" ~ "A: Alimentary tract",
      group == "B" ~ "B: Blood",
      group == "C" ~ "C: Cardiovascular",
      group == "D" ~ "D: Dermatologicals",
      group == "G" ~ "G: Genito-urinary",
      group == "H" ~ "H: Systemic hormonal",
      group == "J" ~ "J: Antiinfectives",
      group == "L" ~ "L: Antineoplastic",
      group == "M" ~ "M: Musculo-skeletal",
      group == "N" ~ "N: Nervous system",
      group == "P" ~ "P: Antiparasitic",
      group == "R" ~ "R: Respiratory",
      group == "S" ~ "S: Sensory organs",
      group == "V" ~ "V: Various",
      TRUE        ~ group
    )
  ) %>%
  ggplot(aes(x = reorder(group_label, n), y = n, fill = n)) +
  geom_col(width = 0.7) +
  geom_text(aes(label = n), hjust = -0.2, size = 3.5) +
  scale_fill_viridis_c(option = "plasma", guide = "none") +
  coord_flip() +
  labs(
    title    = "Level 5 Drug Substances per Anatomical Group",
    subtitle = "WHO ATC/DDD Index",
    x        = NULL,
    y        = "Number of Unique Chemical Substances"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title    = element_text(face = "bold", hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5, color = "grey40")
  )
```

## Hierarchy Depth Within a Branch

```{r branch_depth, fig.width=8, fig.height=5}
# Count codes at each level within Nervous System (N)
atc_codes %>%
  filter(grepl("^N", atc_code)) %>%
  count(level) %>%
  mutate(level_label = case_when(
    level == 1 ~ "1: Anatomical",
    level == 2 ~ "2: Therapeutic",
    level == 3 ~ "3: Pharmacological",
    level == 4 ~ "4: Chemical",
    level == 5 ~ "5: Substance"
  )) %>%
  ggplot(aes(x = factor(level), y = n, fill = factor(level))) +
  geom_col(width = 0.6) +
  geom_text(aes(label = n), vjust = -0.5, size = 4) +
  scale_fill_brewer(palette = "Blues", guide = "none") +
  scale_x_discrete(labels = c(
    "1:\nAnatomical", "2:\nTherapeutic", "3:\nPharmacological",
    "4:\nChemical", "5:\nSubstance"
  )) +
  labs(
    title    = "Codes by Level Within Nervous System (N)",
    subtitle = "Exponential growth from group to individual substances",
    x        = "ATC Level",
    y        = "Number of Codes"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title    = element_text(face = "bold", hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5, color = "grey40")
  )
```

---

# Practical Recipes

## Recipe 1: Find Every Drug in a Therapeutic Subgroup

```{r recipe1, eval=FALSE}
# All Level 5 drugs under proton-pump inhibitors (A02BC)
atc_crawl(roots = "A02BC")$codes %>%
  filter(nchar(atc_code) == 7) %>%
  select(atc_code, atc_name)
```

## Recipe 2: Build a Complete Class-to-Drug Map

```{r recipe2}
# Map: every Level 4 chemical class → all Level 5 substances in it
class_to_drug <- atc_codes %>%
  filter(level == 5) %>%
  mutate(chemical_class = substr(atc_code, 1, 5)) %>%
  select(chemical_class, substance_code = atc_code, substance_name = atc_name)

# Which chemical classes contain "fluoroquinolone"?
atc_codes %>%
  filter(level == 4, grepl("fluoroquinolone", atc_name, ignore.case = TRUE)) %>%
  inner_join(class_to_drug, by = c("atc_code" = "chemical_class")) %>%
  select(chemical_class = atc_code, class_name = atc_name,
         substance_code, substance_name)
```

## Recipe 3: Check If a Drug Belongs to a Class

```{r recipe3}
drug_in_class <- function(code, class_prefix) {
  grepl(paste0("^", class_prefix), code)
}

# Is atorvastatin a cardiovascular drug?
drug_in_class("C10AA05", "C")  # TRUE

# Is it an antiinfective?
drug_in_class("C10AA05", "J")  # FALSE
```

---

# Key Takeaways

1. **Every ATC code is a node in a strict 5-level tree** — parent-child
   relationships are deterministic from the code structure.

2. **Use `get_atc_hierarchy()` for live data** and `atc_level()` / `atc_parent()`
   for offline computation on cached data.

3. **Level 4 (chemical class) is the sweet spot** for pharmacoepidemiology —
   it groups drugs by mechanism, which is clinically meaningful.

4. **Level 5 (substance) is where DDD values live** — DDDs are assigned to
   individual substances, not to classes.

---

# Session Information

```{r session_info}
sessionInfo()
```
