---
title: "Computing Defined Daily Doses from Prescription Data"
author: "Lucas VHH TRAN"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Computing Defined Daily Doses from Prescription Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width  = 7,
  fig.height = 5,
  message    = FALSE,
  warning    = FALSE
)
```

# Introduction

You have prescription records — a pharmacy database, a claims file, or
electronic health record (EHR) extracts — and you need to answer a
pharmacoepidemiology question: *how much drug was consumed?*

The challenge is that prescriptions come in many forms:

- **Different strengths** — 20 mg vs. 40 mg tablets of the same drug
- **Different quantities** — 30 tablets for one patient, 90 for another
- **Different units** — milligrams, grams, micrograms, international units
- **Different routes** — oral, parenteral, rectal, transdermal

The **Defined Daily Dose (DDD)** solves this. By expressing every prescription
in a common unit — *the assumed average maintenance dose per day for an adult* —
you can meaningfully aggregate and compare drug utilisation.

This vignette walks you through the practical workflow:

1. Computing DDDs from raw prescription records
2. Aggregating to DDD per 1000 inhabitants per day (DID)
3. Handling different administration routes
4. Dealing with drugs that lack DDDs
5. Building a complete analysis pipeline

```{r load_package}
library(atcddd)
library(dplyr)
```

---

# Setup and Data Loading

The package bundles a WHO ATC/DDD snapshot that works completely offline.
We load it to inspect the reference data before computing.

```{r load_data}
codes_path <- system.file("extdata", "WHO_ATC_codes_2026-07-14.csv",
                          package = "atcddd")
ddd_path   <- system.file("extdata", "WHO_ATC_DDD_2026-07-14.csv",
                          package = "atcddd")

atc_codes <- readr::read_csv(codes_path, show_col_types = FALSE)
atc_ddd   <- readr::read_csv(ddd_path, show_col_types = FALSE)

cat(sprintf("Bundled data: %d ATC codes, %d DDD entries",
            nrow(atc_codes), nrow(atc_ddd)))
```

> ⚠️ All code in this vignette evaluates offline — no internet connection
> is required.

---

# Step-by-Step: Computing DDDs for a Single Drug

Let's start simple: one drug, one prescription.

A patient receives **acetylsalicylic acid (aspirin)** as 500 mg tablets,
100 tablets dispensed.

```{r single_drug}
aspirin_rx <- data.frame(
  atc_code      = "N02BA01",
  quantity      = 100,
  strength      = 500,
  strength_unit = "mg",
  adm_r         = "O"
)

compute_ddd(aspirin_rx)
```

The output adds four columns to your data:

| Column | Value | Meaning |
|--------|-------|---------|
| `total_amount` | 50000 | The total drug amount: `quantity x strength` (in mg) |
| `ddd_value` | 3 | The WHO DDD for oral aspirin (in grams) |
| `ddd_unit` | g | The unit of the DDD |
| `ddd_ratio` | 16.7 | How many DDDs this prescription represents |

**Interpretation**: 100 tablets of 500 mg aspirin = 50 g of drug substance.
The WHO DDD for oral aspirin is 3 g, so this prescription corresponds to
16.7 DDDs (`50 g / 3 g = 16.7`).

The function automatically converts units — it noticed the strength was in
milligrams while the DDD was in grams and handled the conversion.

---

# Multi-Drug Prescription Data

Real-world data has multiple patients and multiple drugs. Let's create a
realistic scenario with 5 prescription records for 3 patients.

```{r multi_drug}
prescriptions <- data.frame(
  patient_id    = c(1, 1, 2, 2, 3),
  atc_code      = c("N02BA01", "C10AA05", "N02BA01", "A10BA02", "C07AB03"),
  drug_name     = c("aspirin", "atorvastatin", "aspirin", "metformin", "atenolol"),
  quantity      = c(100, 30, 60, 90, 28),
  strength      = c(500, 20, 500, 500, 50),
  strength_unit = c("mg", "mg", "mg", "mg", "mg"),
  adm_r         = c("O", "O", "O", "O", "O")
)

ddd_result <- compute_ddd(prescriptions)
ddd_result
```

Each row now shows the DDD ratio alongside the original prescription data.

**Let's interpret the results:**

- **Patient 1** received 16.7 DDDs of aspirin (100 x 500 mg / 3000 mg per DDD)
  **plus** 30 DDDs of atorvastatin (30 x 20 mg / 20 mg per DDD) — a total of
  46.7 DDDs across two drugs.

- **Patient 2** received 10 DDDs of aspirin and 22.5 DDDs of metformin
  (90 x 500 mg = 45 g; DDD = 2 g; 45 / 2 = 22.5).

- **Patient 3** received 18.7 DDDs of atenolol
  (28 x 50 mg = 1400 mg; DDD = 75 mg; 1400 / 75 = 18.7).

## Per-Patient Summary

Group by `patient_id` to see total drug utilisation per patient:

```{r per_patient}
ddd_result %>%
  group_by(patient_id) %>%
  summarise(
    n_prescriptions = n(),
    total_ddd       = sum(ddd_ratio, na.rm = TRUE),
    .groups = "drop"
  )
```

---

# DDD Per 1000 Inhabitants Per Day (DID)

The DID is the standard population-level metric in drug utilisation research.
It answers: *on an average day, how many people out of 1000 receive one DDD?*

$$\text{DID} = \frac{\text{Total DDDs}}{\text{Population} \times \text{Days}} \times 1000$$

```{r did}
compute_did(ddd_result, population = 500, days = 30)
```

**Interpretation**: A DID of 0.65 means that, on an average day during the
30-day study period, approximately 0.65 out of every 1000 inhabitants
(or about 1 in 1500) received one DDD of the studied medications.

DID values let you compare drug utilisation across:

- **Regions** — is antibiotic use higher in country A or B?
- **Time periods** — did statin use increase after a policy change?
- **Drug classes** — which therapeutic class has the highest utilisation?

> ⚠️ DID is a **population** metric, not a clinical one. A DID of 30 for
> statins tells you that 3% of the population uses a statin DDD each day —
> it does not tell you whether any individual patient is on the right dose.

---

# Route-Specific DDDs

The same drug can have **different DDD values** for different administration
routes. The `ddd_route_comparison()` function shows all route-specific DDDs
for a given drug.

## Paracetamol Route Comparison

```{r route_comparison}
ddd_route_comparison("N02BE01")
```

Paracetamol has the same DDD (3 g) across oral, parenteral, and rectal
routes in this data snapshot. Many other drugs do differ by route — let's
examine one.

## Aspirin: Where Route Really Matters

```{r aspirin_routes}
ddd_route_comparison("N02BA01")
```

Oral and rectal aspirin have a DDD of **3 g**, but parenteral (IV) aspirin
has a DDD of just **1 g**. This makes sense — IV administration bypasses
first-pass metabolism, so less drug is needed for the same effect.

## Mixed Routes in Practice

If your prescription data contains multiple administration routes, you must
specify `adm_r` so the correct DDD is used.

```{r mixed_routes}
mixed_route_rx <- data.frame(
  patient_id    = c(1, 2),
  drug_name     = c("aspirin", "aspirin (IV)"),
  atc_code      = c("N02BA01", "N02BA01"),
  quantity      = c(30, 30),
  strength      = c(500, 500),
  strength_unit = c("mg", "mg"),
  adm_r         = c("O", "P")
)

compute_ddd(mixed_route_rx)
```

Same drug. Same quantity. Different route:

| Route | Total amount | DDD | DDD ratio |
|-------|-------------|-----|-----------|
| Oral (O) | 15 g | 3 g | **5.0** |
| Parenteral (P) | 15 g | 1 g | **15.0** |

If you had ignored the route and assumed oral for both, the IV prescription
would be miscalculated by a factor of three.

---

# Handling Missing DDDs

Not every drug has a WHO DDD assignment. This is by design — the WHO
deliberately withholds DDDs for certain categories (topicals, combinations,
vaccines, etc.).

## DDD Coverage in the Bundled Data

The `ddd_availability()` function gives you an immediate overview of
which anatomical groups have good DDD coverage and which do not.

```{r availability}
ddd_availability()
```

**Key observations:**

- Groups **J** (antiinfectives), **N** (nervous system), and **C**
  (cardiovascular) have high DDD coverage — these are systemic drugs
  where DDD is well-defined.
- Groups **D** (dermatologicals) and **S** (sensory organs) have low
  coverage — topicals and ophthalmics depend on body surface area or
  condition severity, making a standard daily dose impractical.

## What Happens When You Try to Compute DDDs for a Drug Without One

`compute_ddd()` will warn you and return `NA` in the `ddd_ratio` column.

```{r missing_ddd, warning=TRUE}
missing_rx <- data.frame(
  atc_code      = c("N02BA01", "A01AB08"),
  drug_name     = c("aspirin", "neomycin"),
  quantity      = c(100, 30),
  strength      = c(500, 100),
  strength_unit = c("mg", "mg"),
  adm_r         = c("O", "O")
)

result_missing <- compute_ddd(missing_rx)
result_missing
```

Neomycin (`A01AB08`) has no DDD — it is a topical antibiotic, and the WHO
does not assign DDDs for topical preparations. The `ddd_ratio` is `NA`.

## What to Do About Missing DDDs

1. **Document the gap** — report which drugs in your dataset lack DDDs
2. **Calculate coverage** — use `ddd_availability()` to show the percentage
3. **Proceed with caution** — exclude drugs without DDDs from DDD-based
   analyses and note this in your limitations section
4. **Consider alternatives** — for very low coverage datasets, raw
   `total_amount` may be a more honest metric

```{r coverage_check}
# Count drugs with and without DDDs in your dataset
result_missing %>%
  summarise(
    total      = n(),
    with_ddd   = sum(!is.na(ddd_ratio)),
    pct_covered = round(100 * mean(!is.na(ddd_ratio)), 1)
  )
```

---

# Complete Analysis Pipeline

Now let's bring everything together. The typical pharmacoepidemiology
workflow is:

> **Drug names → ATC codes → DDDs → DID**

## Step 1: Resolve Drug Names to ATC Codes

Use `resolve_batch()` to convert a list of drug names into ATC codes,
working entirely offline against the bundled database.

```{r pipeline_resolve}
drug_names <- c("aspirin", "atorvastatin", "metformin", "atenolol")
resolved <- resolve_batch(drug_names, source = "local")
resolved
```

The result includes the ATC code, official name, and the oral-route DDD
for each drug.

## Step 2: Build Prescription Records

```{r pipeline_rx}
rx <- data.frame(
  drug_name     = drug_names,
  quantity      = c(100, 30, 90, 28),
  strength      = c(500, 20, 500, 50),
  strength_unit = c("mg", "mg", "mg", "mg")
)
```

## Step 3: Join, Compute DDDs, and Compute DID

Pipe the entire workflow together with `%>%`:

```{r pipeline_full, warning=FALSE}
rx %>%
  left_join(resolved, by = c("drug_name" = "query")) %>%
  compute_ddd() %>%
  compute_did(population = 500, days = 30)
```

**What just happened:**

1. `left_join()` merged each drug name with its resolved ATC code and DDD info
2. `compute_ddd()` calculated the DDD ratio for each prescription row
3. `compute_did()` aggregated across all rows to produce the population metric

The pipeline scales to hundreds of drugs and thousands of prescription
records — the only limit is your computer's memory.

---

# Best Practices

### Always Check DDD Coverage

Before running any analysis, check how many of your drugs have DDDs.
Low coverage can invalidate your conclusions.

```{r best_coverage}
ddd_availability() %>%
  filter(pct_with_ddd < 50) %>%
  select(anatomical_group, group_name, pct_with_ddd)
```

### Match on Administration Route

The same drug can have different DDDs for different routes. Always include
`adm_r` in your prescription data. When the route is unknown, defaulting to
oral (`adm_r = "O"`) is the safest assumption — most DDDs are defined for
the oral route.

### Document Missing DDDs

> ⚠️ **Transparency is more important than perfect coverage.**

For every analysis, report:
- How many unique drugs in your dataset lacked DDDs
- Which anatomical groups they belonged to
- Whether this could bias your results

### DID Is for Populations, Not Patients

DID measures population-level drug utilisation — it tells you about
consumption patterns across thousands of people, not about individual
prescribing. Use it for:

- **Trend analysis** — is utilisation increasing over time?
- **Cross-regional comparisons** — does region A use more antibiotics than B?
- **Policy evaluation** — did a reimbursement change affect utilisation?

Do **not** use DID to judge the appropriateness of individual prescriptions.

### Use the Same Data Snapshot for Reproducibility

The package bundles a fixed WHO snapshot (`2026-07-14`). For published
research, always cite the snapshot date so others can reproduce your
results exactly.

---

# Key Takeaways

1. **`compute_ddd()` converts raw prescription data to DDDs** — it handles
   unit conversion, route matching, and missing DDDs automatically.

2. **`compute_did()` normalises to per-1000-inhabitants-per-day** — the
   standard metric for cross-population comparisons.

3. **Route matters** — always include `adm_r` in your data. The same drug
   can yield very different DDD counts for different routes.

4. **Missing DDDs are expected** — use `ddd_availability()` to check coverage
   and document gaps transparently.

5. **`resolve_batch()` bridges drug names and ATC codes** — the complete
   pipeline from clinical names to published metrics runs in a few lines of code.

---

# Session Information

```{r session_info}
sessionInfo()
```
