---
title: "Working with Defined Daily Doses (DDD)"
author: "Lucas VHH TRAN"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    includes:
      in_header: header.html
vignette: >
  %\VignetteIndexEntry{Working with Defined Daily Doses (DDD)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width  = 8,
  fig.height = 5,
  message    = FALSE,
  warning    = FALSE
)
```

```{r include-lightbox, echo=FALSE, results='asis'}
knitr::asis_output(paste(readLines('lightbox.html'), collapse='\n'))
```

# Introduction

The **Defined Daily Dose (DDD)** is the cornerstone of quantitative drug
utilisation research. It is the assumed average maintenance dose per day for
a drug used for its main indication in adults.

> ⚠️ **The DDD is a technical unit of measurement — not a clinical recommendation.**
> It enables comparisons across drugs, regions, and time, but does not tell you
> what to prescribe for an individual patient.

This vignette covers:

- Understanding DDD values and why many are missing
- Analysing DDD coverage across drug classes
- Computing DDDs from real prescription data
- Comparing DDDs across administration routes
- Common pitfalls and how to avoid them

```{r load_package}
library(atcddd)
library(dplyr)
library(ggplot2)
library(tidyr)
library(stringr)
```

---

# Loading the Data

```{r load_data}
codes_path <- system.file("extdata", "WHO_ATC_codes_2026-07-14.csv",
                          package = "atcddd")
ddd_path   <- system.file("extdata", "WHO_ATC_DDD_2026-07-14.csv",
                          package = "atcddd")

atc_codes <- readr::read_csv(codes_path, show_col_types = FALSE)
atc_ddd   <- readr::read_csv(ddd_path, show_col_types = FALSE)

cat(sprintf("Loaded %d DDD entries across %d unique substances",
            nrow(atc_ddd),
            n_distinct(atc_ddd$atc_code)))
```

---

# Understanding the DDD Table

The DDD table has a specific structure:

| Column | Description | Example |
|--------|-------------|---------|
| `source_code` | The Level 4 parent page from which the DDD was parsed | `N02BE` |
| `atc_code` | The Level 5 substance code | `N02BE01` |
| `atc_name` | Substance name | `paracetamol` |
| `ddd` | Defined Daily Dose value | `3` |
| `uom` | Unit of measurement | `g` |
| `adm_r` | Route of administration | `O` (oral) |
| `note` | Additional WHO notes | `` |

The **same substance** can have **multiple DDD entries** — one for each
route of administration:

```{r multiple_routes}
# Drugs with multiple administration routes
atc_ddd %>%
  count(atc_code, sort = TRUE) %>%
  filter(n > 1) %>%
  head(10) %>%
  left_join(distinct(atc_ddd, atc_code, atc_name), by = "atc_code") %>%
  select(atc_code, atc_name, n_routes = n)
```

Paracetamol appears with 4 routes — oral (O), parenteral (P), rectal (R),
and others — each with its own DDD value.

---

# DDD Availability: What to Expect

## By Anatomical Group

```{r ddd_by_group, fig.width=10, fig.height=6}
atc_ddd %>%
  filter(!is.na(adm_r), adm_r != "NA", adm_r != "") %>%
  mutate(anatomical_group = substr(atc_code, 1, 1)) %>%
  group_by(anatomical_group) %>%
  summarise(
    total_entries  = n(),
    with_ddd       = sum(!is.na(ddd) & ddd != "NA"),
    without_ddd    = sum(is.na(ddd) | ddd == "NA"),
    pct_with_ddd   = round(100 * with_ddd / total_entries, 1),
    .groups = "drop"
  ) %>%
  arrange(pct_with_ddd)
```

## Visualising DDD Coverage

```{r ddd_coverage_viz, fig.width=10, fig.height=6}
atc_ddd %>%
  filter(!is.na(adm_r), adm_r != "NA", adm_r != "") %>%
  mutate(
    anatomical_group = substr(atc_code, 1, 1),
    has_ddd = !is.na(ddd) & ddd != "NA"
  ) %>%
  count(anatomical_group, has_ddd) %>%
  group_by(anatomical_group) %>%
  mutate(pct = round(100 * n / sum(n), 1)) %>%
  ungroup() %>%
  filter(has_ddd) %>%
  ggplot(aes(x = reorder(anatomical_group, pct), y = pct, fill = pct)) +
  geom_col(width = 0.7) +
  geom_text(aes(label = paste0(pct, "%")), hjust = -0.2, size = 3.5) +
  scale_fill_viridis_c(option = "mako", direction = -1, guide = "none") +
  coord_flip(ylim = c(0, 100)) +
  labs(
    title    = "DDD Coverage by Anatomical Group",
    subtitle = "Percentage of substance-route entries with an assigned DDD",
    x        = "ATC Anatomical Group",
    y        = "% with DDD Value"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title    = element_text(face = "bold", hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5, color = "grey40")
  )
```

**Interpretation**:

- **High coverage (>70%)**: Systemic drugs — antiinfectives (J), cardiovascular (C),
  nervous system (N). These have well-defined daily doses.
- **Low coverage (<30%)**: Topicals (D), sensory organs (S) — dosing is
  patient-specific and depends on treated area or condition severity.

---

# Why DDDs Are Missing

The WHO deliberately **withholds** DDDs for specific categories. This is not
a data quality issue — it is by design.

```{r missing_ddd_summary}
# Categorise the nature of missing DDDs
atc_ddd %>%
  mutate(has_ddd = !is.na(ddd) & ddd != "NA") %>%
  count(has_ddd) %>%
  mutate(pct = round(100 * n / sum(n), 1))
```

### Categories That Typically Lack DDDs

| Category | Example Codes | Reason |
|----------|--------------|--------|
| **Topical preparations** | D01-D07 | Variable absorption; dose depends on body surface area |
| **Fixed-dose combinations** | Codes ending in digits (e.g., C09BA02) | WHO policy: no DDD for combination products |
| **Ophthalmics / otics** | S01-S03 | Local application, minimal systemic absorption |
| **Vaccines** | J07 | Prophylactic, not daily dosing |
| **Contrast media** | V08 | Diagnostic use, not therapeutic |
| **Herbal / alternative** | Various | Lack of standardised active ingredient quantities |

---

# Computing DDDs from Prescription Data

This is the core pharmacoepidemiology use case: you have prescription records
with drug codes, quantities, and strengths, and you need DDDs per patient.

## The Formula

$$\text{Number of DDDs} = \frac{\text{Quantity} \times \text{Strength per unit}}{\text{DDD value (in same unit)}}$$

## Worked Example

```{r compute_ddd_example}
# Example: prescription records for 3 patients
prescriptions <- tibble(
  patient_id = c(1, 1, 2, 3, 3),
  atc_code   = c("N02BE01", "C10AA05", "N02BE01", "A02BC01", "C10AA05"),
  quantity   = c(30, 30, 60, 28, 90),       # tablets/capsules
  strength   = c(500, 20, 500, 20, 40),      # mg per unit
  unit       = c("mg", "mg", "mg", "mg", "mg")
)

# Join with DDD data (using the oral route, adm_r = "O")
ddd_ref <- atc_ddd %>%
  filter(!is.na(ddd), ddd != "NA", adm_r == "O") %>%
  select(atc_code, atc_name, ddd, uom) %>%
  mutate(ddd_value = as.numeric(ddd))

prescriptions %>%
  left_join(ddd_ref, by = "atc_code") %>%
  mutate(
    # Convert DDD to mg if needed
    ddd_mg = case_when(
      uom == "g"  ~ ddd_value * 1000,
      uom == "mg" ~ ddd_value,
      uom == "mcg" ~ ddd_value / 1000,
      TRUE        ~ NA_real_
    ),
    total_mg   = quantity * strength,
    n_ddd      = total_mg / ddd_mg
  ) %>%
  select(patient_id, atc_code, atc_name, quantity, strength,
         total_mg, ddd_mg, n_ddd)
```

## Aggregate by Patient

```{r patient_ddd}
prescriptions %>%
  left_join(
    atc_ddd %>%
      filter(!is.na(ddd), ddd != "NA", adm_r == "O") %>%
      mutate(ddd_value = as.numeric(ddd)) %>%
      select(atc_code, atc_name, ddd_value, uom),
    by = "atc_code"
  ) %>%
  mutate(
    ddd_mg = case_when(
      uom == "g"   ~ ddd_value * 1000,
      uom == "mg"  ~ ddd_value,
      uom == "mcg" ~ ddd_value / 1000,
      TRUE         ~ NA_real_
    ),
    total_mg = quantity * strength,
    n_ddd    = total_mg / ddd_mg
  ) %>%
  group_by(patient_id) %>%
  summarise(
    n_drugs    = n(),
    total_ddds = sum(n_ddd, na.rm = TRUE),
    .groups    = "drop"
  )
```

---

# DDD per 1000 Inhabitants per Day (DID)

The standard population-level metric in drug utilisation research:

$$\text{DID} = \frac{\text{Total DDDs consumed in period}}{\text{Population} \times \text{Days in period}} \times 1000$$

```{r did_example}
# Hypothetical: national paracetamol consumption over 1 year
total_paracetamol_ddds <- 850000000  # 850 million DDDs
population             <- 68000000    # 68 million inhabitants
days_in_year           <- 365

did <- (total_paracetamol_ddds / (population * days_in_year)) * 1000
cat(sprintf("Paracetamol DID: %.1f DDD per 1000 inhabitants per day", did))
```

A DID of 34.3 means that, on any given day, approximately 34 out of every 1000
people in the population use one DDD of paracetamol.

---

# DDD by Route of Administration

Different routes often have different DDDs for the same drug:

```{r ddd_by_route}
atc_ddd %>%
  filter(!is.na(ddd), ddd != "NA", !is.na(adm_r), adm_r != "NA", adm_r != "") %>%
  mutate(
    adm_label = case_when(
      adm_r == "O"    ~ "Oral",
      adm_r == "P"    ~ "Parenteral",
      adm_r == "R"    ~ "Rectal",
      adm_r == "N"    ~ "Nasal",
      adm_r == "SL"   ~ "Sublingual",
      adm_r == "TD"   ~ "Transdermal",
      adm_r == "Inhal" ~ "Inhalation",
      adm_r == "V"    ~ "Vaginal",
      TRUE            ~ adm_r
    )
  ) %>%
  count(adm_label, sort = TRUE) %>%
  mutate(pct = round(100 * n / sum(n), 1)) %>%
  head(10)
```

Oral (O) and parenteral (P) routes dominate the DDD assignments — reflecting
the systemic drugs for which DDDs are most useful.

---

# Cross-Referencing DDDs With Chemical Classes

## Average DDD by Therapeutic Class

```{r avg_ddd_by_class}
# Join DDD data with Level 4 classification
atc_ddd %>%
  filter(!is.na(ddd), ddd != "NA", adm_r == "O") %>%
  mutate(
    ddd_value  = as.numeric(ddd),
    class_code = substr(atc_code, 1, 5)   # Level 4 chemical class
  ) %>%
  group_by(class_code) %>%
  summarise(
    n_drugs        = n_distinct(atc_code),
    mean_ddd       = mean(ddd_value, na.rm = TRUE),
    median_ddd     = median(ddd_value, na.rm = TRUE),
    min_ddd        = min(ddd_value, na.rm = TRUE),
    max_ddd        = max(ddd_value, na.rm = TRUE),
    .groups        = "drop"
  ) %>%
  left_join(
    atc_codes %>% filter(nchar(atc_code) == 5) %>% select(atc_code, class_name = atc_name),
    by = c("class_code" = "atc_code")
  ) %>%
  filter(n_drugs >= 3) %>%
  arrange(desc(mean_ddd)) %>%
  select(class_code, class_name, n_drugs, mean_ddd, median_ddd) %>%
  head(10)
```

Large differences in mean DDD across classes reflect real pharmacological
differences — antibiotics need grams per day, while potent drugs like
ACE inhibitors need milligrams.

---

# DDD Unit Distribution

```{r ddd_units}
# What units do DDDs use?
atc_ddd %>%
  filter(!is.na(ddd), ddd != "NA", !is.na(uom), uom != "NA", uom != "") %>%
  count(uom, sort = TRUE) %>%
  mutate(pct = round(100 * n / sum(n), 1))
```

The three most common DDD units are:

| Unit | Typical for | Example |
|------|------------|---------|
| **g** (grams) | Antibiotics, analgesics, old drugs with high doses | amoxicillin: 1.5 g |
| **mg** (milligrams) | Modern drugs with lower effective doses | atorvastatin: 20 mg |
| **mcg** (micrograms) | Highly potent drugs | levothyroxine: 150 mcg |

---

# Common Pitfalls

### 1. Ignoring Route of Administration

```{r pitfall_route}
# Paracetamol has different DDDs for different routes
atc_ddd %>%
  filter(atc_code == "N02BE01", !is.na(ddd), ddd != "NA") %>%
  select(atc_code, atc_name, ddd, uom, adm_r)
```

If you compute DDDs assuming the oral route when patients received IV
paracetamol, your numbers will be wrong. Always match on `adm_r`.

### 2. Unit Mismatches

A DDD of 3 g paired with tablets of 500 mg requires unit conversion:

```r
ddd_g  <- 3
ddd_mg <- ddd_g * 1000  # 3000 mg
tablets_per_ddd <- ddd_mg / 500  # 6 tablets per DDD
```

### 3. Assuming All Drugs Have DDDs

As shown above, ~40-50% of DDD entries are deliberately `NA`. Always check
coverage before reporting — if topical drugs dominate your dataset, DDD-based
metrics may be misleading.

### 4. Using DDD as a Clinical Dose

> The DDD is **not** a recommended or prescribed dose. It is a technical
> measurement unit. Individual patients may need very different doses.

---

# Live DDD Retrieval

For the latest DDD values, use the live API:

```{r live_ddd, eval=FALSE}
# Get current DDD data for a drug from WHO
aspirin_live <- get_atc_data("N02BA01")
aspirin_live %>% select(atc_code, atc_name, ddd, uom, adm_r, note)
```

---

# Key Takeaways

1. **DDDs enable apples-to-apples comparisons** — they normalise drug consumption
   across different strengths, package sizes, and formulations.

2. **Match by route of administration** (`adm_r`) — the same drug can have
   different DDDs for oral, parenteral, and rectal routes.

3. **Convert units carefully** — DDDs come in g, mg, mcg, TU, MU, and other
   units. Always convert to a common unit before computing total DDDs.

4. **Document missing DDDs** — if 40% of your study drugs lack DDDs, report it.
   Transparency about coverage is more important than perfect coverage.

5. **DDD ≠ clinical dose** — use DDD for population-level comparisons, not
   individual patient management.

---

# Session Information

```{r session_info}
sessionInfo()
```
