---
title: "atcddd: Crawling WHO ATC/DDD and Joining External Dictionaries"
author: "Lucas VHHTRAN"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


## ATC Classification:
The ATC system assigns codes to drugs, with each level representing a different level of detail. For example: 

+ Level 1: Main anatomical group (e.g., nervous system) 
+ Level 2: Therapeutic or pharmacological subgroup (e.g., psycholeptics) 
+ Level 3-5: Further subdivisions into chemical, pharmacological, or therapeutic subgroups and chemical substances 







```{r, results = "hide", message=FALSE, warning=FALSE}
library(tidyverse)
library(atcddd)
out_dir <- getwd()
res <- atc_crawl(rate = 0.5, progress = TRUE)
#paths <- atc_write_csv(res, dir = out_dir, stamp = TRUE)
#atc_write_manifest(paths)
```
## ATC
```{r}
res$codes %>% DT::datatable()
```


## DDD
```{r}
res$ddd %>% DT::datatable()
```




