---
title: "Getting Started with atcddd"
author: "Lucas VHH TRAN"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    includes:
      in_header: header.html
vignette: >
  %\VignetteIndexEntry{Getting Started with atcddd}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width  = 7,
  fig.height = 5,
  message    = FALSE,
  warning    = FALSE
)
```

```{r include-lightbox, echo=FALSE, results='asis'}
knitr::asis_output(paste(readLines('lightbox.html'), collapse='\n'))
```

# Introduction

The **atcddd** package provides comprehensive tools for working with the
World Health Organization's **Anatomical Therapeutic Chemical (ATC) Classification
System** and **Defined Daily Dose (DDD)** values. Whether you need to crawl
the latest WHO data, validate ATC codes, navigate the five-level drug hierarchy,
or compute DDD metrics from prescription records — `atcddd` gives you tidy,
reproducible workflows in R.

## What is the ATC Classification System?

The ATC system divides drugs into five levels of increasing specificity:

| Level | Name               | Example code | Example name     |
|-------|--------------------|-------------|------------------|
| 1     | Anatomical main group  | `N`         | Nervous system   |
| 2     | Therapeutic subgroup   | `N02`       | Analgesics       |
| 3     | Pharmacological subgroup | `N02B`    | Other analgesics and antipyretics |
| 4     | Chemical subgroup      | `N02BE`    | Anilides         |
| 5     | Chemical substance     | `N02BE01`  | paracetamol      |

## What is the Defined Daily Dose (DDD)?

The DDD is the assumed average maintenance dose per day for a drug used for its
main indication in adults. It is a unit of measurement, **not** a recommended dose.
DDDs enable researchers to compare drug utilisation across populations, regions,
and time periods — even when drugs come in different strengths and package sizes.

---

# Quick Start

## Loading the Package

```{r load_package}
library(atcddd)
library(dplyr)
library(ggplot2)
```

## The 14 Main Anatomical Groups

The ATC system starts with 14 single-letter groups. Use `atc_roots_default()`
to list them:

```{r roots}
roots <- atc_roots_default()
roots
```

## Validate Any ATC Code

`is_valid_atc_code()` checks whether a string conforms to the official ATC format:

```{r validate}
# Valid codes at each level
is_valid_atc_code("N")          # Level 1 — TRUE
is_valid_atc_code("N02")        # Level 2 — TRUE
is_valid_atc_code("N02BE01")    # Level 5 — TRUE

# Invalid codes
is_valid_atc_code("n02be01")    # lowercase — FALSE
is_valid_atc_code("N02-BE01")   # hyphens  — FALSE
is_valid_atc_code("N02BE011")   # too long — FALSE
is_valid_atc_code("ZZZ")        # bad pattern — FALSE
```

---

# Retrieving ATC Data

## High-Level API (Recommended)

The `get_atc_data()` function provides a clean interface to the WHO database:

```{r api_demo, eval=FALSE}
# Get full data for aspirin
aspirin <- get_atc_data("N02BA01")
aspirin

# Get all analgesics with their sub-classifications
analgesics <- get_atc_data("N02", include_children = TRUE)
head(analgesics, 10)

# Get multiple groups at once
cns_cv <- get_atc_data(c("N", "C"))
cns_cv
```

## Low-Level Crawling (Advanced)

For more control, use `atc_crawl()` directly:

```{r crawl_demo, eval=FALSE}
# Crawl dermatological drugs (ATC group D)
res <- atc_crawl(
  roots     = "D",
  rate      = 0.5,
  progress  = TRUE,
  max_codes = 100
)

# The result is a list with two tibbles
str(res, max.level = 1)

# Codes table: atc_code + atc_name
head(res$codes, 10)

# DDD table: source_code, atc_code, atc_name, ddd, uom, adm_r, note
head(res$ddd, 5)
```

---

# Working with the Hierarchy

The `get_atc_hierarchy()` function returns the full parent-child tree:

```{r hierarchy_demo, eval=FALSE}
# Get complete hierarchy under analgesics (N02)
tree <- get_atc_hierarchy("N02")

tree %>%
  select(atc_code, atc_name, level, parent_code, has_children) %>%
  head(10)
```

Column descriptions:

- **`parent_code`** — the code one level up in the hierarchy (NA for Level 1)
- **`has_children`** — TRUE if this node has sub-classifications below it
- **`level`** — hierarchy depth (1-5)

---

# Understanding DDD Data Quality

**Not all drugs have DDD values — and this is expected.** The WHO does not assign
DDDs to several categories of medicines. Understanding *why* prevents false
conclusions when NA values appear in your data.

```{r ddd_quality_demo, eval=FALSE}
# Systemic antibiotics — most WILL have DDD values
abx <- atc_crawl(roots = "J01AA", rate = 0.5, max_codes = 30)
cat("Systemic antibiotics with DDD:",
    sum(!is.na(abx$ddd$ddd)), "/", nrow(abx$ddd))

# Topical dermatologicals — most WON'T have DDD values
derm <- atc_crawl(roots = "D01", rate = 0.5, max_codes = 50)
cat("Topical dermatologicals with DDD:",
    sum(!is.na(derm$ddd$ddd)), "/", nrow(derm$ddd))
```

### Why NA Values Are Correct

| Drug Category | Reason for Missing DDD |
|---------------|----------------------|
| **Topicals / dermatologicals** | Variable absorption; dosing depends on body surface area |
| **Fixed-dose combinations** | WHO policy: no DDD for combinations |
| **Ophthalmics, otics, nasal preps** | Local application with minimal systemic absorption |
| **Older / rarely used drugs** | Not prioritized for DDD assignment |

---

# Exporting Results

## Save to CSV

```{r export_demo, eval=FALSE}
res <- atc_crawl(roots = "D", max_codes = 50)
paths <- atc_write_csv(res, dir = "output", stamp = TRUE)
# Creates:
#   output/WHO_ATC_codes_2025-07-14.csv
#   output/WHO_ATC_DDD_2025-07-14.csv
```

## Generate Reproducibility Manifests

```{r manifest_demo, eval=FALSE}
# Each manifest includes SHA256 checksums for every file
atc_write_manifest(paths)
# Creates: output/MANIFEST.csv
```

Use manifests to document exactly which version of WHO data your analysis used —
critical for reproducible pharmacoepidemiology.

---

# Offline Data Access

Download the WHO ATC/DDD Index once, then load from cache for offline use:

```{r offline_data}
library(atcddd)

# Download data if not already cached
tryCatch(
  atc_load_db(),
  error = function(e) {
    message("Downloading WHO ATC/DDD Index (one-time setup)...")
    atc_download()
  }
)
db <- atc_load_db()
atc_codes <- db$codes
atc_ddd   <- db$ddd

# Quick look
cat(sprintf("ATC codes table: %d rows\n", nrow(atc_codes)))
cat(sprintf("ATC DDD table:   %d rows\n", nrow(atc_ddd)))

head(atc_codes, 5)
```

## Search Drug Names Offline

```{r offline_search}
# Find all codes containing "paracetamol"
atc_codes %>%
  filter(grepl("paracetamol", atc_name, ignore.case = TRUE))

# Find all Level 2 therapeutic subgroups
atc_codes %>%
  filter(nchar(atc_code) == 3) %>%
  head(10)
```

---

# Analysis Example: DDD Availability by Route

```{r ddd_analysis}
# Analyse DDD patterns across administration routes
atc_ddd %>%
  filter(!is.na(adm_r) & adm_r != "NA" & adm_r != "") %>%
  count(adm_r, sort = TRUE) %>%
  mutate(pct = round(100 * n / sum(n), 1)) %>%
  head(10)
```

---

# Analysis Example: Hierarchy Distribution

```{r hierarchy_viz, fig.width=8, fig.height=5}
# Classify codes by level
atc_codes %>%
  mutate(
    code_len = nchar(atc_code),
    level = case_when(
      code_len == 1 ~ "Level 1\n(Anatomical)",
      code_len == 3 ~ "Level 2\n(Therapeutic)",
      code_len == 4 ~ "Level 3\n(Pharmacological)",
      code_len == 5 ~ "Level 4\n(Chemical)",
      code_len == 7 ~ "Level 5\n(Substance)",
      TRUE         ~ "Other"
    )
  ) %>%
  filter(level != "Other") %>%
  count(level) %>%
  mutate(level = factor(level, levels = c(
    "Level 1\n(Anatomical)", "Level 2\n(Therapeutic)",
    "Level 3\n(Pharmacological)", "Level 4\n(Chemical)",
    "Level 5\n(Substance)"
  ))) %>%
  ggplot(aes(x = level, y = n, fill = level)) +
  geom_col(width = 0.7, alpha = 0.9) +
  geom_text(aes(label = scales::comma(n)), vjust = -0.5, size = 3.5) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title    = "Distribution of ATC Codes by Hierarchy Level",
    subtitle = paste("Bundled WHO data snapshot:", nrow(atc_codes), "total codes"),
    x        = NULL,
    y        = "Number of Codes",
    caption  = "Source: WHO ATC/DDD Index"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "none",
    plot.title      = element_text(face = "bold", hjust = 0.5),
    plot.subtitle   = element_text(hjust = 0.5, color = "grey40"),
    panel.grid.major.x = element_blank()
  )
```

This pyramid shape is characteristic — one Level 1 root produces many Level 5
substances, reflecting how drug classes branch into specific chemical entities.

---

# Best Practices

## Rate Limiting

Always use appropriate delays to respect WHO server resources:

```{r best_rate, eval=FALSE}
# Conservative: 1 request per second
atc_crawl(rate = 1.0)

# Minimum delay for large crawls: 0.5 seconds
atc_crawl(rate = 0.5)
```

## Caching

The package caches HTTP responses automatically via **memoise**. Repeated
requests for the same URL return cached data instantly:

```{r best_cache, eval=FALSE}
# First run: fetches from WHO (slow)
res1 <- atc_crawl(roots = "D", max_codes = 20)

# Second run: reads from cache (fast)
res2 <- atc_crawl(roots = "D", max_codes = 20)
```

Cache location (platform-dependent):
`rappdirs::user_cache_dir("atcddd")`

## Reproducibility

For any published analysis, generate a checksum manifest so reviewers can verify
your data:

```{r best_repro, eval=FALSE}
res   <- atc_crawl(roots = c("C", "N"))
paths <- atc_write_csv(res, dir = "data", stamp = TRUE)
atc_write_manifest(paths)
```

---

# Troubleshooting

| Symptom | Likely Cause | Solution |
|---------|-------------|----------|
| Many `NA` in DDD column | Expected for topicals, combinations, ophthalmics, otics | See *Understanding DDD Data Quality* above |
| Slow crawling | Server latency or low `rate` value | Increase `rate` to 0.5–1.0; use `max_codes` for testing |
| HTTP errors | WHO website down or network issue | Check https://www.whocc.no/atc_ddd_index/ |
| Out of memory | Full crawl too large | Crawl one group at a time with `max_codes` |

---

# Next Steps

- **Navigating the ATC Hierarchy**: `vignette("atc-hierarchy", package = "atcddd")`
- **Working with Defined Daily Doses**: `vignette("ddd-analysis", package = "atcddd")`

---

# Session Information

```{r session_info}
sessionInfo()
```
